---
title: "Bayesian Linear Regression with Spike-and-Slab Priors"
format:
  html:
    toc: true
    code-fold: true
---

## Introduction

This document demonstrates Bayesian Linear Regression using **spike-and-slab priors**, as described in the notes. We implement a Gibbs sampler that updates parameters from their full conditionals, computes posterior summaries (including posterior inclusion probabilities), and evaluates convergence using diagnostic statistics.

## Model Setup and Data Simulation

```{r}
set.seed(123)

# Simulate data
n <- 100
p <- 5
X <- cbind(1, matrix(rnorm(n * p), n, p))
beta_true <- c(2, 1.2, 0, 0, 0, 1.5)  # some zeros (spike)
sigma_true <- 1

y <- X %*% beta_true + rnorm(n, 0, sigma_true)
```

## Gibbs Sampler Implementation

We use the following priors:

- $\alpha_i \mid \sigma_b^2 \sim \mathcal{N}(0, \sigma_b^2)$

- $\delta_i \mid \pi \sim \text{Bernoulli}(\pi)$

- $\pi \sim \text{Beta}(a, b)$

- $\sigma_b^2 \sim S_b \chi^{-2}(v_b)$

- $\sigma^2 \sim S \chi^{-2}(v)$

```{r}
# Hyperparameters
a_pi <- 1
b_pi <- 1
v_b <- 4
S_b <- 1
v <- 4
S <- 1

# Gibbs sampler setup
n_iter <- 4000
burn_in <- 1000
chains <- 2

# Storage
beta_samples <- vector("list", chains)
delta_samples <- vector("list", chains)
pi_samples <- vector("list", chains)
sigma2_samples <- vector("list", chains)
sigma2b_samples <- vector("list", chains)

for (c in 1:chains) {
  # Initial values
  alpha <- rnorm(ncol(X), 0, 1)
  delta <- rbinom(ncol(X), 1, 0.5)
  sigma2 <- 1
  sigma2_b <- 1
  pi <- 0.5
  
  # Containers
  alpha_chain <- matrix(NA, n_iter, ncol(X))
  delta_chain <- matrix(NA, n_iter, ncol(X))
  sigma2_chain <- numeric(n_iter)
  sigma2b_chain <- numeric(n_iter)
  pi_chain <- numeric(n_iter)
  
  for (t in 1:n_iter) {
    # Sample each alpha_i given delta_i
    for (j in 1:ncol(X)) {
      X_j <- X[, j]
      r_j <- y - X %*% (alpha * delta) + X_j * alpha[j] * delta[j]
      
      if (delta[j] == 1) {
        var_j <- sigma2 / (t(X_j) %*% X_j + sigma2 / sigma2_b)
        mean_j <- as.numeric(var_j * t(X_j) %*% r_j / sigma2)
        alpha[j] <- rnorm(1, mean_j, sqrt(var_j))
      } else {
        alpha[j] <- rnorm(1, 0, sqrt(sigma2_b))  # prior draw
      }
    }
    
    # Sample delta_i given alpha_i
    for (j in 1:ncol(X)) {
      X_j <- X[, j]
      r_j <- y - X %*% (alpha * delta) + X_j * alpha[j] * delta[j]
      rss0 <- sum((r_j)^2)
      rss1 <- sum((r_j - X_j * alpha[j])^2)
      
      logodds <- 0.5 / sigma2 * (rss0 - rss1) + log(pi) - log(1 - pi)
      p1 <- 1 / (1 + exp(-logodds))
      delta[j] <- rbinom(1, 1, p1)
    }
    
    # Update pi
    pi <- rbeta(1, a_pi + sum(delta), b_pi + ncol(X) - sum(delta))
    
    # Update sigma_b^2
    p_incl <- sum(delta)
    v_b_tilde <- v_b + p_incl
    S_b_tilde <- (sum((alpha * delta)^2) + v_b * S_b) / v_b_tilde
    sigma2_b <- v_b_tilde * S_b_tilde / rchisq(1, df = v_b_tilde)
    
    # Update sigma^2
    resid <- y - X %*% (alpha * delta)
    v_tilde <- v + n
    S_tilde <- (sum(resid^2) + v * S) / v_tilde
    sigma2 <- v_tilde * S_tilde / rchisq(1, df = v_tilde)
    
    # Store
    alpha_chain[t, ] <- alpha * delta
    delta_chain[t, ] <- delta
    pi_chain[t] <- pi
    sigma2_chain[t] <- sigma2
    sigma2b_chain[t] <- sigma2_b
  }
  
  beta_samples[[c]] <- alpha_chain
  delta_samples[[c]] <- delta_chain
  pi_samples[[c]] <- pi_chain
  sigma2_samples[[c]] <- sigma2_chain
  sigma2b_samples[[c]] <- sigma2b_chain
}
```

## Posterior Summaries

```{r}
posterior_summary <- function(samples, probs = c(0.025, 0.5, 0.975)) {
  c(mean = mean(samples), sd = sd(samples), quantile(samples, probs = probs))
}

# Combine chains
beta_all <- do.call(rbind, beta_samples)
pi_all <- unlist(pi_samples)
delta_all <- do.call(rbind, delta_samples)

beta_summary <- t(apply(beta_all[burn_in:nrow(beta_all), ], 2, posterior_summary))
rownames(beta_summary) <- paste0("beta", 0:p)

PIP <- colMeans(delta_all[burn_in:nrow(delta_all), ])

round(beta_summary, 4)
cat("\nPosterior Inclusion Probabilities (PIP):\n")
round(PIP, 3)
```

## Trace Plots

```{r}
par(mfrow = c(3, 2))
for (j in 1:ncol(X)) {
  plot(beta_samples[[1]][, j], type = "l", main = paste("Trace: beta", j - 1, "(chain 1)"),
       xlab = "Iteration", ylab = expression(beta))
  abline(h = beta_true[j], col = "red", lwd = 2, lty = 2)
}
```

## Autocorrelation Plots

```{r}
par(mfrow = c(3, 2))
for (j in 1:ncol(X)) {
  acf(beta_samples[[1]][burn_in:n_iter, j], main = paste("ACF: beta", j - 1))
}
```

## Convergence Diagnostics

```{r}
convergence_stats <- function(samples) {
  n <- length(samples)
  ac1 <- cor(samples[-1], samples[-n])
  mcse <- sd(samples) * sqrt((1 + ac1) / n)
  a <- floor(0.1 * n); b <- floor(0.5 * n)
  z <- (mean(samples[1:a]) - mean(samples[(n - b + 1):n])) /
       sqrt(var(samples[1:a]) / a + var(samples[(n - b + 1):n]) / b)
  ess <- n / (1 + 2 * sum(acf(samples, plot = FALSE)$acf[-1]))
  c(autocorr1 = ac1, mcse = mcse, geweke_z = z, ess = ess)
}

# Apply to chain 1
conv_results <- t(apply(beta_samples[[1]][burn_in:n_iter, ], 2, convergence_stats))
rownames(conv_results) <- paste0("beta", 0:p)
round(conv_results, 4)
```

### Gelman–Rubin R-hat

```{r}
Rhat <- function(ch1, ch2) {
  n <- nrow(ch1)
  m <- 2
  chain_means <- c(colMeans(ch1), colMeans(ch2))
  overall_mean <- colMeans(rbind(ch1, ch2))
  B <- n * apply(rbind(colMeans(ch1), colMeans(ch2)), 2, var)
  W <- (apply(ch1, 2, var) + apply(ch2, 2, var)) / 2
  var_hat <- ((n - 1) / n) * W + (1 / n) * B
  sqrt(var_hat / W)
}

Rhat_values <- Rhat(beta_samples[[1]][burn_in:n_iter, ], beta_samples[[2]][burn_in:n_iter, ])
round(Rhat_values, 3)
```

## Posterior Mean vs True Values

```{r}
par(mar = c(5, 5, 4, 2))
plot(beta_true, beta_summary[, "mean"], pch = 19, col = "blue",
     xlab = "True Coefficients", ylab = "Posterior Mean Estimates",
     main = "Posterior Mean vs True Coefficients")
abline(0, 1, col = "red", lwd = 2, lty = 2)
legend("topleft", legend = c("Posterior Means", "y = x line"), 
       col = c("blue", "red"), pch = c(19, NA), lty = c(NA, 2))
```
