---
title: "Bayesian Fine-Mapping with BayesC"
author:
  - name: "<br>Peter Sørensen"
    affiliation:
      - "<br>Center for Quantitative Genetics and Genomics, Aarhus University, Denmark"
  - name: "<br>Palle Duun Rohde"
    affiliation:
      - "<br>Genomic Medicine, Department of Health Science and Technology, Aalborg University, Denmark"
format:
  revealjs:
    theme: simple
    transition: fade
    slide-number: true
    toc: false
    incremental: false
    auto-scale: true
    embed-resources: true
    self-contained-math: true
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-svg-full.js"
        
    width: 1366
    height: 768
    margin: 0.05

    math:
      scale: 0.9

    css: ../slides.css
  beamer:
    theme: default
    slide-level: 2
    keep-tex: true
    pdf-engine: lualatex
    fontsize: 11pt
    header-includes: |
      \setbeamertemplate{itemize item}{\textendash}
      \setbeamertemplate{itemize subitem}{\textbullet}
      \setbeamertemplate{itemize subsubitem}{$\cdot$}
      \makeatletter
      \patchcmd{\beamer@framesetup}{\centering}{}{}{}
      \makeatother

execute:
  echo: false
  warning: false
  message: false
---


## GWAS Regions Rarely Identify a Unique Causal Variant

- A GWAS association usually identifies an **associated region**, not a unique causal variant.

- Because nearby SNPs are correlated through **linkage disequilibrium (LD)**, several SNPs can show strong association even when only one is causal.

Fine-mapping moves from

$$
\text{associated region}
\quad\longrightarrow\quad
\text{small set of plausible causal variants}.
$$

The goal is to quantify which variants remain plausible causal explanations after accounting for LD.


## Fine-Mapping Connects Association Signals to Biology

- **Biological interpretation:** connect associated variants to genes, regulatory elements, and molecular mechanisms.
- **Functional follow-up:** prioritize variants, genes, and pathways for experimental studies.
- **Quantifying uncertainty:** distinguish one strongly supported variant from several LD-linked alternatives.
- **Functional information:** combine genetic evidence with eQTL, chromatin, conservation, and coding annotations.



## Standard GWAS Tests One SNP at a Time

In a standard **genome-wide association study (GWAS)**, each SNP is tested separately for association with the phenotype.


For SNP $j$, fit a separate regression:

$$
y_i
=
\mu
+
x_{ij}\beta_j
+
e_i,
\qquad
e_i \sim \mathcal{N}(0,\sigma^2).
$$

Test whether the SNP is associated with the phenotype:

$$
H_0:\beta_j=0
\qquad\text{versus}\qquad
H_1:\beta_j\neq0.
$$

Each SNP is tested **one at a time**.

## Standard GWAS Reports Marginal Evidence

For every SNP, GWAS typically reports:

- estimated effect $\widehat{\beta}_j$
- standard error $\operatorname{SE}(\widehat{\beta}_j)$
- $z$- or $t$-statistic
- $p$-value

The result is a marker-by-marker summary of **marginal association evidence**.


**Important:** GWAS identifies **associated SNPs**, not necessarily the causal SNP.


## LD Makes Multiple SNPs Look Associated

Nearby SNPs are often correlated because of **linkage disequilibrium (LD)**. If SNP $j^\ast$ is causal,

$$
y=x_{j^\ast}\beta_{j^\ast}+e.
$$

Another SNP $j$ can still show strong marginal association when

$$
\operatorname{Cor}(x_j,x_{j^\ast})\neq0
\qquad\text{but}\qquad
\beta_j=0.
$$

A GWAS peak can therefore represent an **associated region**, not a unique causal variant.




```{r}
#| include: false
# Same simulated 30-SNP locus used in the tutorial

geno <- read.delim(
  "../data/genotypes.txt",
  header = TRUE,
  row.names = 1,
  check.names = FALSE
)

X_all <- as.matrix(geno)
region <- 114:143
X <- scale(
  X_all[, region, drop = FALSE],
  center = TRUE,
  scale = TRUE
)

n <- nrow(X)
m <- ncol(X)
snp <- colnames(X)
R <- cor(X)

set.seed(123)
causal <- 11
beta_true <- rep(0, m)
beta_true[causal] <- 0.35

g <- drop(X %*% beta_true)
h2 <- 0.30
vare <- var(g) * (1 - h2) / h2
y <- g + rnorm(n, 0, sqrt(vare))

# Marginal GWAS in the same locus

gwas_p <- numeric(m)
for (j in seq_len(m)) {
  fit_j <- lm(y ~ X[, j])
  gwas_p[j] <- summary(fit_j)$coefficients[2, 4]
}

# Base-R two-state BayesC Gibbs sampler used in the tutorial
fit_bayesc <- function(
  X,
  y,
  niter = 2000,
  pi = 0.03,
  start_h2 = 0.30
) {
  n <- nrow(X)
  m <- ncol(X)

  stopifnot(
    pi > 0,
    pi < 1,
    start_h2 > 0,
    start_h2 < 1
  )

  vary <- var(y)
  varg <- vary * start_h2
  vare <- vary * (1 - start_h2)
  varb <- varg / (m * pi)

  nub <- 4
  nue <- 4
  scale_b <- (nub - 2) / nub * varb
  scale_e <- (nue - 2) / nue * vare

  beta <- numeric(m)
  delta <- integer(m)
  mu <- mean(y)
  ycorr <- y - mu
  xpx <- colSums(X^2)

  beta_mcmc <- matrix(0, nrow = niter, ncol = m)
  delta_mcmc <- matrix(0L, nrow = niter, ncol = m)

  for (iter in seq_len(niter)) {
    ycorr <- ycorr + mu
    mu <- rnorm(1, mean = mean(ycorr), sd = sqrt(vare / n))
    ycorr <- ycorr - mu

    for (j in seq_len(m)) {
      xj <- X[, j]
      old_beta <- beta[j]
      residual_without_j <- ycorr + xj * old_beta
      rhs <- sum(xj * residual_without_j)

      Cj <- xpx[j] + vare / varb
      post_mean <- rhs / Cj
      post_var <- vare / Cj

      log_prob_zero <- log(1 - pi)
      log_prob_active <-
        log(pi) +
        0.5 * (
          log(vare) -
          log(varb) -
          log(Cj) +
          rhs^2 / (vare * Cj)
        )

      max_log <- max(log_prob_zero, log_prob_active)
      prob <- exp(c(log_prob_zero, log_prob_active) - max_log)
      prob <- prob / sum(prob)

      delta[j] <- sample(c(0L, 1L), size = 1, prob = prob)
      beta[j] <- if (delta[j] == 0L) {
        0
      } else {
        rnorm(1, mean = post_mean, sd = sqrt(post_var))
      }

      ycorr <- residual_without_j - xj * beta[j]
    }

    active <- which(delta == 1L)
    nactive <- length(active)
    shape_b <- 0.5 * (nub + nactive)
    rate_b <- 0.5 * nub * scale_b
    if (nactive > 0) {
      rate_b <- rate_b + 0.5 * sum(beta[active]^2)
    }
    varb <- 1 / rgamma(1, shape = shape_b, rate = rate_b)

    resid <- y - mu - drop(X %*% beta)
    shape_e <- 0.5 * (nue + n)
    rate_e <- 0.5 * (nue * scale_e + sum(resid^2))
    vare <- 1 / rgamma(1, shape = shape_e, rate = rate_e)

    beta_mcmc[iter, ] <- beta
    delta_mcmc[iter, ] <- delta
  }

  colnames(beta_mcmc) <- colnames(X)
  colnames(delta_mcmc) <- colnames(X)

  list(beta = beta_mcmc, delta = delta_mcmc)
}

set.seed(1)
fit_C <- fit_bayesc(
  X,
  y,
  niter = 2000,
  pi = 0.03,
  start_h2 = 0.30
)

keep <- 501:2000
PIP_C <- colMeans(fit_C$delta[keep, , drop = FALSE])
beta_C <- colMeans(fit_C$beta[keep, , drop = FALSE])

bayesc_results <- data.frame(
  SNP = snp,
  marker = seq_len(m),
  LD_with_causal = R[causal, ],
  true_effect = beta_true,
  PIP = PIP_C,
  posterior_mean_effect = beta_C,
  causal = seq_len(m) == causal
)

noncausal_index <- setdiff(seq_len(m), causal)
best_noncausal_index <- noncausal_index[which.max(PIP_C[noncausal_index])]
best_noncausal <- PIP_C[best_noncausal_index]
causal_rank <- rank(-PIP_C, ties.method = "min")[causal]

bayesc_summary <- data.frame(
  quantity = c(
    "Causal SNP PIP",
    "Causal SNP rank",
    "Best non-causal PIP",
    "PIP gap",
    "Posterior expected active SNPs",
    "Best competitor marker",
    "Best competitor SNP"
  ),
  value = c(
    round(PIP_C[causal], 3),
    causal_rank,
    round(best_noncausal, 3),
    round(PIP_C[causal] - best_noncausal, 3),
    round(sum(PIP_C), 3),
    best_noncausal_index,
    snp[best_noncausal_index]
  )
)

bayesc_top <- bayesc_results[order(-bayesc_results$PIP), c(
  "marker",
  "SNP",
  "PIP",
  "LD_with_causal",
  "posterior_mean_effect",
  "causal"
)]

bayesc_top$PIP <- round(bayesc_top$PIP, 3)
bayesc_top$LD_with_causal <- round(bayesc_top$LD_with_causal, 3)
bayesc_top$posterior_mean_effect <- round(bayesc_top$posterior_mean_effect, 3)

result_interpretation <- data.frame(
  question = c(
    "Does the causal SNP rank first?",
    "How much support does the strongest competitor have?",
    "How much larger is the causal PIP?",
    "How many active SNPs does the posterior expect?"
  ),
  answer = c(
    ifelse(causal_rank == 1, "Yes", "No"),
    paste0(
      "PIP = ",
      round(best_noncausal, 3),
      " at marker ",
      best_noncausal_index
    ),
    paste0(
      "PIP gap = ",
      round(PIP_C[causal] - best_noncausal, 3)
    ),
    paste0(
      "sum(PIPs) = ",
      round(sum(PIP_C), 3)
    )
  )
)


focal_marker <- which.max(PIP_C)
local_ld_threshold <- 0.5
local_candidates <- which((R[focal_marker, ]^2) >= local_ld_threshold)
local_order <- local_candidates[order(PIP_C[local_candidates], decreasing = TRUE)]
local_cum_pip <- cumsum(PIP_C[local_order])
lcs_cut <- which(local_cum_pip >= 0.90)[1]
if (is.na(lcs_cut)) {
  lcs_cut <- length(local_order)
  warning(
    "Cumulative PIP in the local LD neighborhood does not reach 0.90."
  )
}
local_credible_set <- local_order[seq_len(lcs_cut)]

lcs_table <- data.frame(
  marker = local_order,
  SNP = snp[local_order],
  PIP = round(PIP_C[local_order], 3),
  r2_with_focal = round(R[focal_marker, local_order]^2, 3),
  cumulative_PIP = round(local_cum_pip, 3),
  in_LCS = seq_along(local_order) <= lcs_cut,
  causal = local_order == causal
)

lcs_display <- head(lcs_table, 8)

lcs_summary <- data.frame(
  quantity = c(
    "Top-PIP focal marker",
    "Local LD threshold",
    "Candidate SNPs in local region",
    "0.9 LCS size",
    "Causal SNP in 0.9 LCS?",
    "Causal SNP PIP",
    "Final cumulative PIP"
  ),
  value = c(
    focal_marker,
    paste0("r^2 >= ", local_ld_threshold),
    length(local_candidates),
    length(local_credible_set),
    ifelse(causal %in% local_credible_set, "Yes", "No"),
    round(PIP_C[causal], 3),
    round(local_cum_pip[lcs_cut], 3)
  )
)
```


## LD Explains the Cluster of Marginal Associations

The same simulated 30-SNP region is shown in both panels. The causal SNP is marked in red.

::: {.columns}

::: {.column width="50%"}

#### LD with the causal SNP

```{r}
#| fig-width: 5
#| fig-height: 4
par(mar = c(4, 4, 1, 1))
plot(
  seq_len(m),
  R[causal, ]^2,
  type = "h",
  lwd = 5,
  ylim = c(0, 1),
  xlab = "SNP index",
  ylab = expression(r^2)
)
points(causal, 1, pch = 19, col = "red")
abline(v = causal, col = "red", lty = 2)
```

:::

::: {.column width="50%"}

#### Marginal GWAS evidence

```{r}
#| fig-width: 5
#| fig-height: 4
par(mar = c(4, 4, 1, 1))
plot(
  seq_len(m),
  -log10(gwas_p),
  type = "h",
  lwd = 5,
  xlab = "SNP index",
  ylab = expression(-log[10](p))
)
points(
  causal,
  -log10(gwas_p[causal]),
  pch = 19,
  col = "red"
)
abline(v = causal, col = "red", lty = 2)
```

:::

:::


## Fine-Mapping: Key Inputs

Fine-mapping aims to identify which SNPs within a GWAS-associated region remain plausible causal variants after accounting for LD.


- **Marker-level association evidence:** e.g. marginal effect estimates $\hat{\beta}_j$ and their standard errors $\mathrm{SE}(\hat{\beta}_j)$.

- **LD between candidate SNPs:** represented by the LD correlation matrix, e.g. $R=\operatorname{cor}(X)$.

- **A sparse joint model for genetic effects:** e.g. BayesC / spike-and-slab.

Here we use individual-level genotypes and phenotypes. Summary-statistic methods instead combine compatible GWAS evidence with external LD.

## Fine-Mapping: Key Outputs

- **Posterior inclusion probabilities (PIPs):** probability that each SNP has a non-zero effect.

- **Posterior effect estimates:** model-averaged SNP effects.

- **Prioritized variant sets:** small sets of plausible causal variants.

- **Fine-mapping uncertainty:** one leading SNP or several LD-linked alternatives.


## Fine-Mapping Fits SNP Effects Jointly


All SNPs are fitted **jointly**, allowing correlated SNPs to compete to explain the same association signal:

$$
y_i
=
\mu
+
\sum_{j=1}^{m}x_{ij}b_j
+
e_i,
\qquad
e_i\sim\mathcal{N}(0,\sigma_e^2).
$$

::: {.columns}

::: {.column width="50%"}

- $y$ — phenotype
- $\mu$ — overall mean
- $X$ — genotype matrix

:::

::: {.column width="50%"}

- $b$ — SNP effects
- $e$ — residual effects
- $\sigma_e^2$ — residual variance


:::

:::


## BayesC Uses Inclusion Indicators

BayesC assumes that each SNP is either **inactive** or **active**. The point mass at zero creates a **sparse model** in which many SNP effects are exactly zero.


Introduce an inclusion indicator:

$$
d_j \in \{0,1\},
\qquad
d_j \sim \operatorname{Bernoulli}(\pi).
$$

Conditional on the indicator:

$$
b_j\mid(d_j=0)=0,
$$

$$
b_j\mid(d_j=1)
\sim
\mathcal{N}(0,\sigma_b^2).
$$

## BayesC Is a Spike-and-Slab Prior

Equivalently,

$$
b_j
\sim
(1-\pi)\delta_0
+
\pi\mathcal{N}(0,\sigma_b^2).
$$

```{r}
#| fig-width: 7
#| fig-height: 3.2
par(mar = c(3, 3, 1, 1))
x_prior <- seq(-4, 4, length.out = 400)
plot(
  x_prior,
  dnorm(x_prior),
  type = "l",
  lwd = 3,
  ylim = c(0, 0.68),
  axes = FALSE,
  xlab = expression(b[j]),
  ylab = "Prior density"
)
axis(1, at = c(-4, 0, 4), labels = c("", "0", ""))
segments(0, 0, 0, 0.58, lwd = 6)
text(0, 0.62, "spike", cex = 0.9)
text(1.8, 0.20, "slab", cex = 0.9)
```

$\pi$ controls how often SNPs are active; $\sigma_b^2$ controls the variance of non-zero effects.


## Correlated SNPs Compete Across Posterior Models

The sampler repeatedly explores possible explanations of the association signal.

::: {.columns}

::: {.column width="43%"}

#### Marker state at iteration $t$

$$
b_j^{(t)}=
\begin{cases}
0, & d_j^{(t)}=0,\\
\text{non-zero draw}, & d_j^{(t)}=1.
\end{cases}
$$

The sampler updates $d_j$ and $b_j$ conditional on the current states of the other SNPs.

:::

::: {.column width="57%"}

#### Sampled inclusion states

```{r}
#| fig-width: 6
#| fig-height: 4
show_iterations <- unique(round(seq(
  min(keep),
  max(keep),
  length.out = 60
)))
delta_show <- fit_C$delta[show_iterations, , drop = FALSE]

par(mar = c(4, 4, 1, 1))
image(
  seq_len(m),
  seq_len(nrow(delta_show)),
  t(delta_show),
  col = c("grey95", "#2C3E50"),
  xlab = "SNP index",
  ylab = "Posterior sample",
  axes = FALSE
)
axis(1)
axis(2)
abline(v = causal, col = "red", lty = 2, lwd = 2)
```

:::

:::


## Posterior Samples Become Marker-Level Evidence

::: {.columns}

::: {.column width="50%"}

#### Posterior inclusion probability

$$
\begin{aligned}
\operatorname{PIP}_j
&=\Pr(d_j=1\mid\mathcal D), \\
\widehat{\operatorname{PIP}}_j
&=\frac{1}{T}\sum_{t=1}^{T}d_j^{(t)}.
\end{aligned}
$$

**PIP:** how often SNP $j$ is needed in posterior models.

:::

::: {.column width="50%"}

#### Posterior mean effect

$$
\begin{aligned}
\widehat b_j
&=E(b_j\mid\mathcal D), \\
\widehat b_j
&=\frac{1}{T}\sum_{t=1}^{T}b_j^{(t)}.
\end{aligned}
$$

**Posterior mean:** the effect after averaging across posterior models.

:::

:::


## Joint Modelling Redistributes the GWAS Signal

The marginal GWAS signal is compared with support from the joint BayesC model.

::: {.columns}

::: {.column width="50%"}

#### Marginal GWAS

```{r}
#| fig-width: 5
#| fig-height: 4
par(mar = c(4, 4, 1, 1))
plot(
  seq_len(m),
  -log10(gwas_p),
  type = "h",
  lwd = 5,
  xlab = "SNP index",
  ylab = expression(-log[10](p))
)
points(
  causal,
  -log10(gwas_p[causal]),
  pch = 19,
  col = "red"
)
abline(v = causal, col = "red", lty = 2)
```

:::

::: {.column width="50%"}

#### BayesC PIPs

```{r}
#| fig-width: 5
#| fig-height: 4
par(mar = c(4, 4, 1, 1))
plot(
  seq_len(m),
  PIP_C,
  type = "h",
  lwd = 5,
  ylim = c(0, 1),
  xlab = "SNP index",
  ylab = "Posterior inclusion probability"
)
points(causal, PIP_C[causal], pch = 19, col = "red")
abline(v = causal, col = "red", lty = 2)
```

:::

:::


## PIP and Effect Size Answer Different Questions

::: {.columns}

::: {.column width="50%"}

#### Posterior inclusion probability

$$
\operatorname{PIP}_j
=
\Pr(b_j\neq0\mid\mathcal D).
$$

Measures how often SNP $j$ is included in posterior models.

A SNP can have a **high PIP but a moderate effect** if it is frequently active with a relatively small effect.

:::

::: {.column width="50%"}

#### Posterior mean effect

$$
E(b_j\mid\mathcal D).
$$

Measures the model-averaged effect of SNP $j$, including iterations in which its effect is zero.

A SNP can have a **lower PIP but large non-zero effects** when it is selected.

:::

:::

## Constructing a Local 0.9 Credible Set

A practical **local credible set (LCS)** summarizes the leading signal as a small set of plausible causal variants.

1. Identify the SNP with the highest PIP.
2. Select its LD neighborhood, for example $r^2\ge0.5$.
3. Rank these SNPs by decreasing PIP.
4. Add their PIPs until

   $$
   \sum_{j\in\mathcal C}\operatorname{PIP}_j
   \ge 0.90.
   $$

## Interpreting the Local Credible Set

- PIPs come from the **joint model** and account for LD.
- A small LCS indicates a well-localized signal.
- A larger LCS indicates uncertainty among correlated SNPs.
- Raw PIPs are used without normalization.

Because BayesC allows multiple active SNPs, this is a **practical local summary**, not a formal $90\%$ posterior coverage set.



## Fine-Mapping Result: Local 0.9 Credible Set

::: {.columns}

::: {.column width="52%"}

#### Variants in the LCS

```{r}
knitr::kable(
  lcs_table[
    seq_len(lcs_cut),
    c(
      "marker",
      "SNP",
      "PIP",
      "r2_with_focal",
      "cumulative_PIP",
      "causal"
    )
  ],
  row.names = FALSE,
  align = "lrrrrl"
)
```

:::

::: {.column width="48%"}

#### Cumulative PIP

```{r}
plot(
  seq_len(nrow(lcs_table)),
  lcs_table$cumulative_PIP,
  type = "b",
  pch = 19,
  xlab = "PIP rank",
  ylab = "Cumulative PIP",
  ylim = c(0, 1),
  main = ""
)

abline(h = 0.90, lty = 2, col = "red")
abline(v = lcs_cut, lty = 2, col = "blue")
```

:::

:::

The LCS contains `r length(local_credible_set)` SNPs and includes the simulated causal SNP: **`r ifelse(causal %in% local_credible_set, "yes", "no")`**.


## Functional Information Can Refine Fine-Mapping

When LD leaves several plausible SNPs, functional information can refine their prior support.

::: {.columns}

::: {.column width="46%"}

#### Functional information

Examples include:

- coding consequences;
- regulatory and eQTL evidence;
- chromatin accessibility;
- conservation and tissue- or cell-specific annotations.

:::

::: {.column width="8%"}
:::

::: {.column width="46%"}

#### Bayesian interpretation

Bayesian updating combines:

- genetic evidence from the locus;
- prior information for each SNP.

It changes a SNP's **prior plausibility**. It does not declare that SNP to be causal.

:::

:::




## Annotation-Informed Inclusion Probabilities

Standard BayesC gives every SNP the same prior probability of being active:

$$
d_j \sim \operatorname{Bernoulli}(\pi).
$$

Annotations can make that probability **SNP-specific**:

$$
P(d_j=1)=\pi_j.
$$

A relevant annotation may therefore raise a SNP's prior probability of having a non-zero effect.


## Annotation-Informed Effect Variances

Standard BayesC uses a common variance for active effects:

$$
b_j\mid d_j=1 \sim \mathcal{N}(0,\sigma_b^2).
$$

Annotations can make the active-effect variance **SNP-specific**:

$$
b_j\mid d_j=1 \sim \mathcal{N}(0,\sigma_{b,j}^2).
$$

An annotated SNP may therefore be assigned a larger prior variance for its non-zero effect.



## Annotations Reweight Competing Causal Explanations

Suppose two SNPs have similar statistical evidence because they are in strong LD.

::: {.columns}

::: {.column width="48%"}

#### Equal versus annotation-informed priors

$$
\begin{aligned}
P(d_1=1)=P(d_2=1)&=\pi
&&\text{(equal priors)}, \\
P(d_1=1)=\pi_1,\quad P(d_2=1)&=\pi_2
&&\text{(annotation-informed)}.
\end{aligned}
$$

where a relevant annotation can imply $\pi_1>\pi_2$.

:::

::: {.column width="52%"}

#### Schematic prior reweighting

```{r}
#| fig-width: 5
#| fig-height: 3.2
prior_weights <- matrix(
  c(1, 1, 1.5, 0.5),
  nrow = 2
)

par(mar = c(4, 4, 1, 1))
barplot(
  prior_weights,
  beside = TRUE,
  names.arg = c("Equal prior", "Annotation-informed"),
  col = c("#C7CEDB", "#536A8A"),
  border = NA,
  yaxt = "n",
  ylab = "Relative prior weight",
  legend.text = c("SNP 1", "SNP 2"),
  args.legend = list(bty = "n", x = "topright")
)
```

:::

:::





## Annotation-Informed Fine-Mapping Workflow

::: {.columns}

::: {.column width="47%"}

#### Standard prior

- GWAS and LD evidence
- A common prior across SNPs
- Posterior PIPs

:::

::: {.column width="6%"}
:::

::: {.column width="47%"}

#### Annotation-informed prior

- GWAS and LD evidence
- SNP-specific prior information
- Annotation-informed PIPs

:::

:::

Both approaches report PIPs, ranked candidates, posterior effects, and local credible sets.

Only the prior assigned to each SNP changes—through its inclusion probability, non-zero effect variance, or both.


## Summary

::: {.columns}

::: {.column width="46%"}

#### From association to joint evidence

- GWAS peaks identify **associated regions**, not unique causal variants.
- Joint fitting makes correlated variants compete to explain the signal.
- BayesC represents SNPs as inactive or active.

:::

::: {.column width="8%"}
:::

::: {.column width="46%"}

#### Interpreting the posterior

- PIPs and posterior means answer different questions.
- A BayesC LCS is practical, not automatically a formal $90\%$ coverage set.
- Functional annotations modify priors, not the genetic evidence.

:::

:::


## Further Reading {.smaller}

[**Companion tutorial: Bayesian Fine-Mapping with BayesC**](../tutorials/bayesian_fine_mapping_bayesc.html)

- **Genome-wide fine-mapping — SBayesRC:** **Wu Y, Zheng Z, Thibaut L, Lin T, Feng Q, Cheng H, Yengo L, Goddard ME, Wray NR, Visscher PM, Zeng J.** Genome-wide fine-mapping improves identification of causal variants. *Nature Genetics* **58**, 940–951 (2026). DOI: [10.1038/s41588-026-02549-3](https://doi.org/10.1038/s41588-026-02549-3)

- **BayesC and BayesR for fine-mapping:** **Shrestha M, Bai Z, Gholipourshahraki T, Hjelholt A, Rohde P, Fuglsang MK, Sørensen P.** Enhanced genetic fine mapping accuracy with Bayesian Linear Regression models in diverse genetic architectures. *PLOS Genetics* **21**(7), e1011783 (2025). DOI: [10.1371/journal.pgen.1011783](https://doi.org/10.1371/journal.pgen.1011783)

- **Original SBayesRC method and prediction study:** **Zheng Z et al.** Leveraging functional genomic annotations and genome coverage to improve polygenic prediction of complex traits within and between ancestries. *Nature Genetics* **56**, 767–777 (2024). DOI: [10.1038/s41588-024-01704-y](https://doi.org/10.1038/s41588-024-01704-y)
